---
title: "An minimal example for Martin"
date: "`r Sys.Date()`"
author: Jitao David Zhang <jitao_david.zhang@roche.com>
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

## Background

In this document, I develop a minimal example to explain my thoughts about testing potential enrichment of a collection of genesets in a small set of genes.


```{r}
library(ribiosUtils)
library(ribiosPlot)
library(knitr)
```

## A list of genes with statistics

Suppose we have twelve genes with given log2 fold changes (logFC) between condition one and condition two. For the sake of simplicity, the genes are sorted by the logFC value in the descending order.

```{r inputData, fig.width=3, fig.height=4}
set.seed(100)
df <- data.frame(Gene=sprintf("Gene%2d",1:12),
                 logFC=sort(rnorm(12, 0, 3), decreasing=TRUE))
biosHeatmap(matrix(df$logFC), dendrogram="none", Rowv=FALSE, Colv=FALSE, labRow=df$Gene, lwid=c(1,1), lhei=c(1,3),
            col="royalbluered",
            cellnote=matrix(round(df$logFC,2)), notecol="orange", color.key.title = "logFC")
```

## Three genesets

And now we suppose we have three genesets, A, B, and C, with following compositions:

* geneset A consists of mostly positively regulated genes;
* geneset B consists of randomly selected genes;
* geneset C consists of mostly negatively regulated genes.

```{r gs, fig.width=3, fig.height=4}
genesetA <- c(rep(1,6), rep(0,6))
genesetB <- rep(0, 12); genesetB[sample(1:12, 6)] <- 1
genesetC <- c(rep(0,6), rep(1,6))
genesets <- cbind(genesetA, genesetB, genesetC)
rownames(genesets) <- df$Gene
biosHeatmap(genesets, dendrogram="none", Rowv=FALSE, Colv=FALSE, labRow=df$Gene, lwid=c(1,1), lhei=c(1,3),
            col="blackyellow", colsep = 1:(ncol(genesets)-1),
            color.key.title = "Membership")
```

## Statistical tests

Now, for all genesets, we can apply Student's t-test, or Wilcoxon's t-test, to determine whether the geneset is enriched of either positively regulated or negatively regulated genes.

The statistical tests has the null hypothesis that *the statistic associated with the genes that belong to the pathway have the same mean (median) value as the values associated with the genes that do not belong to the pathway*. The alternative hypothesis is that *the statistic associated with the genes that belong to the pathway have a different mean (median) value as the values associated with the genes that do not belong to the pathway*.

We report the results of the statistic tests in **Enrichment Scores**, which are defined as absolute log10-transformed p-values multiplied by the sign of difference. If the score is positive, then the geneset is enriched of positively regulated genes; otherwise it is enriched of negatively regulated genes. A score larger than 2 corresponds to an enrichment of positively regulated genes with p-value smaller than 0.01, and a score smaller than -2 corresponds to an enrichment of negatively regulated genes with p-value smaller than 0.01.

```{r enrich}
gsTstat <- function(stat, membership) {
  statPos <- stat[membership==1]
  statNeg <- stat[membership==0]
  tRes <- t.test(statPos, statNeg)
  es <- pScore(tRes$p.value, sign(tRes$statistic))
  return(es)
}
gsWilstat <- function(stat, membership) {
  statPos <- stat[membership==1]
  statNeg <- stat[membership==0]
  wRes <- wilcox.test(statPos, statNeg)
  es <- pScore(wRes$p.value, sign(median(statPos) - median(statNeg)))
  return(es)
}
tRes <- apply(genesets, 2, function(gsMem) gsTstat(df$logFC, gsMem))
wRes <- apply(genesets, 2, function(gsMem) gsWilstat(df$logFC, gsMem))
resDf <- data.frame(TstatResult=tRes,
                    WilconResult=wRes)
kable(resDf)
```

As expected, geneset A and C are enriched of positively and negatively regulated genes, respectively, while B is not significantly enriched of either type (assuming that the statistical significance is set as 0.01).

## Conclusion

I hope this minimal example explains my idea.
